The hydra and the rooster

### 2020-08-08 latex/chapter-schutte.tex
corrections on pdf file

\begin{Coqsrc}
Parameter Ord : Type.
Parameter lt : relation Ord.
Notation "a < b" := (lt a b): schutte_scope.

@@

\begin{Coqsrc}
Axiom AX2 : forall X: Ensemble Ord, 
   (exists a,  (forall y, In X y -> y < a)) ->
   countable X.

Axiom AX3 : forall X : Ensemble Ord,
              countable X -> 
              exists a,  forall y, In X y -> y < a.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma zero_le (alpha : Ord) :  zero <= alpha.
Proof.
  unfold zero, the_least, the; apply iota_ind.
\end{Coqsrc}

@@

subgoal 2 (ID 156) is:
 forall a : Ord, unique (least_member lt ordinal) a -> 
                a <= alpha
\end{Coqanswer}

@@

Since we won't be able to prove the proposition
\linebreak \Verb|{exists! a: Ord, least_member (Empty_set Ord) a|, the only properties we would be able to prove about \texttt{bottom} would be \emph{trivial} properties, 
\emph{i.e.}, satisfied by \emph{any} element of type \texttt{Ord}, like for instance
\texttt{bottom = bottom}, or \texttt{zero <= bottom}.

\begin{Coqbad}
Lemma le_zero_bottom : zero <= bottom. 
Proof. apply zero_le. Qed.

@@

Lemma le_bottom_zero : bottom <= zero.
Proof.
   unfold bottom, the_least, the; apply iota_ind.
\end{Coqbad}

@@

subgoal 2 (ID 414) is:
 forall a : Ord, unique (least_member lt (Empty_set Ord)) a -> 
      a <= zero
\end{Coqanswer}

@@

\begin{Coqsrc}
Definition succ (alpha : Ord) := the_least (fun beta => alpha < beta).
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition succ_spec (alpha:Ord) :=
  least_member   lt (fun z => alpha < z).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma lt_succ (alpha : Ord) :  alpha < succ alpha.
Proof.
  destruct  (succ_ok  alpha);  tauto.
Qed.

@@

Lemma lt_succ_le (alpha beta : Ord):
  alpha < beta -> succ alpha <= beta.
Proof with eauto with schutte.
  intros  H;  pattern (succ alpha); apply the_least_ok ... 
  exists (succ alpha); red;apply lt_succ ...
Qed.
\end{Coqsrc}

\begin{Coqsrc}
Lemma lt_succ_le_2 (alpha beta : Ord):
  alpha < succ beta -> alpha <= beta.

Lemma succ_mono (alpha beta : Ord):
  alpha < beta -> succ alpha < succ beta.

Lemma succ_monoR (alpha beta : Ord) :
 succ alpha < succ beta -> alpha < beta.

Lemma lt_succ_lt (alpha beta : Ord) :
  is_limit beta ->  alpha < beta -> succ alpha < beta.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma finite_lt_omega : forall i: nat,  i < omega.

Lemma lt_omega_finite alpha : Ord) : 
  alpha < omega ->  exists i:nat, alpha =  i.

@@

\begin{Coqsrc}
 Definition segment (A: Ensemble Ord) :=
  forall alpha beta, In A alpha -> beta < alpha -> In A  beta.

@@

\begin{Coqsrc}
Definition ordering_function (f : Ord -> Ord)(A B : Ensemble Ord) :=
 segment A /\
 (forall a, In A a -> In B (f a)) /\
 (forall b, B b -> exists a, In A a /\ f a = b) /\
 forall a b, In A a -> In A b -> a < b ->  f a < f b.

@@

\begin{Coqsrc}
Theorem ordering_le : forall f A B,
    ordering_function f A B ->
    forall alpha, In A alpha -> alpha  <= f alpha.

@@

Lemma le_plus_l (alpha beta : Ord) : alpha <= alpha + beta.

Lemma le_plus_r (alpha beta : Ord) :  beta <= alpha + beta.

Lemma plus_mono_r (alpha beta gamma : Ord) : 
    beta < gamma -> alpha + beta < alpha + gamma.

@@

Lemma finite_plus_ge_omega (n : nat) (alpha : Ord) :
  omega <= alpha -> n + alpha = alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition AP : Ensemble Ord :=
  fun alpha => 
  zero < alpha /\
  (forall beta, beta < alpha ->  beta + alpha = alpha).
\end{Coqsrc}

@@

Lemma omega_second_AP :
  least_member   lt 
                  (fun alpha => 1 < alpha /\ In AP alpha)
                  omega.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma AP_plus_closed (alpha beta gamma : Ord): 
     In AP alpha -> beta < alpha -> gamma < alpha -> beta + gamma < alpha.

@@

Lemma phi0_mono (alpha beta : Ord) :
  alpha < beta ->  phi0 alpha < phi0 beta.

@@

Lemma is_limit_phi0 (alpha : Ord) :
  zero < alpha ->  is_limit (phi0 alpha).

@@

Lemma phi0_le (alpha : Ord) : alpha <= phi0 alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma lt_phi0 (alpha : Ord):
  alpha < epsilon0 -> alpha < phi0 alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition Cr_fun : forall alpha : Ord,
       (forall beta : Ord, beta < alpha -> Ensemble Ord) ->
        Ensemble Ord 
:= 
   fun (alpha :Ord)
        (Cr : forall beta, 
                beta < alpha -> Ensemble Ord) 
        (x : Ord) => (
       (alpha = zero /\ AP x) \/
       (zero < alpha /\
        forall beta (H:beta < alpha),
          the_ordering_segment (Cr beta H) x /\ ord (Cr  beta H) x = x)).

@@

Definition sorted (l: cnf_t) :=
  LocallySorted (fun alpha beta => beta <= alpha) l.

@@

\begin{Coqsrc}
Lemma cnf_lt_epsilon0 : 
 forall l alpha, 
   is_cnf_of alpha l ->  alpha < epsilon0 ->
   Forall (fun beta =>  beta < alpha) l.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma cnf_lt_epsilon0_iff : 
 forall l alpha, 
   is_cnf_of alpha l ->  
   (alpha < epsilon0 <->  Forall (fun beta =>  beta < alpha) l).
\end{Coqsrc}

@@

Theorem inject_mono (beta gamma : T1) :
  T1.lt  beta gamma -> 
  T1.nf beta -> T1.nf gamma -> 
  inject beta < inject gamma.

@@

\begin{Coqsrc}
 Theorem inject_lt_epsilon0 (alpha : T1):
      inject alpha < epsilon0.

>>>>

\begin{Coqsrc}
Parameter Ord : Type.
Parameter lt : relation Ord.
Notation "a o< b" := (lt a b): schutte_scope.

@@

\begin{Coqsrc}
Axiom AX2 : forall X: Ensemble Ord, 
   (exists a,  (forall y, In X y -> y o< a)) ->
   countable X.

Axiom AX3 : forall X : Ensemble Ord,
              countable X -> 
              exists a,  forall y, In X y -> y o< a.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma zero_le (alpha : Ord) :  zero o<= alpha.
Proof.
  unfold zero, the_least, the; apply iota_ind.
\end{Coqsrc}

@@

subgoal 2 (ID 156) is:
 forall a : Ord, unique (least_member lt ordinal) a -> 
                a o<= alpha
\end{Coqanswer}

@@

Since we won't be able to prove the proposition
\linebreak \Verb|{exists! a: Ord, least_member (Empty_set Ord) a|, the only properties we would be able to prove about \texttt{bottom} would be \emph{trivial} properties, 
\emph{i.e.}, satisfied by \emph{any} element of type \texttt{Ord}, like for instance
\texttt{bottom = bottom}, or \texttt{zero o<= bottom}.

\begin{Coqbad}
Lemma le_zero_bottom : zero o<= bottom. 
Proof. apply zero_le. Qed.

@@

Lemma le_bottom_zero : bottom o<= zero.
Proof.
   unfold bottom, the_least, the; apply iota_ind.
\end{Coqbad}

@@

subgoal 2 (ID 414) is:
 forall a : Ord, unique (least_member lt (Empty_set Ord)) a -> 
      a o<= zero
\end{Coqanswer}

@@

\begin{Coqsrc}
Definition succ (alpha : Ord) := the_least (fun beta => alpha o< beta).
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition succ_spec (alpha:Ord) :=
  least_member   lt (fun z => alpha o< z).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma lt_succ (alpha : Ord) :  alpha o< succ alpha.
Proof.
  destruct  (succ_ok  alpha);  tauto.
Qed.

@@

Lemma lt_succ_le (alpha beta : Ord):
  alpha o< beta -> succ alpha o<= beta.
Proof with eauto with schutte.
  intros  H;  pattern (succ alpha); apply the_least_ok ... 
  exists (succ alpha); red;apply lt_succ ...
Qed.
\end{Coqsrc}

\begin{Coqsrc}
Lemma lt_succ_le_2 (alpha beta : Ord):
  alpha o< succ beta -> alpha o<= beta.

Lemma succ_mono (alpha beta : Ord):
  alpha o< beta -> succ alpha o< succ beta.

Lemma succ_monoR (alpha beta : Ord) :
 succ alpha o< succ beta -> alpha o< beta.

Lemma lt_succ_lt (alpha beta : Ord) :
  is_limit beta ->  alpha o< beta -> succ alpha o< beta.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma finite_lt_omega : forall i: nat,  i o< omega.

Lemma lt_omega_finite alpha : Ord) : 
  alpha o< omega ->  exists i:nat, alpha =  i.

@@

\begin{Coqsrc}
 Definition segment (A: Ensemble Ord) :=
  forall alpha beta, In A alpha -> beta o< alpha -> In A  beta.

@@

\begin{Coqsrc}
Definition ordering_function (f : Ord -> Ord)(A B : Ensemble Ord) :=
 segment A /\
 (forall a, In A a -> In B (f a)) /\
 (forall b, B b -> exists a, In A a /\ f a = b) /\
 forall a b, In A a -> In A b -> a o< b ->  f a o< f b.

@@

\begin{Coqsrc}
Theorem ordering_le : forall f A B,
    ordering_function f A B ->
    forall alpha, In A alpha -> alpha  o<= f alpha.

@@

Lemma le_plus_l (alpha beta : Ord) : alpha o<= alpha + beta.

Lemma le_plus_r (alpha beta : Ord) :  beta o<= alpha + beta.

Lemma plus_mono_r (alpha beta gamma : Ord) : 
    beta o< gamma -> alpha + beta o< alpha + gamma.

@@

Lemma finite_plus_ge_omega (n : nat) (alpha : Ord) :
  omega o<= alpha -> n + alpha = alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition AP : Ensemble Ord :=
  fun alpha => 
  zero o< alpha /\
  (forall beta, beta o< alpha ->  beta + alpha = alpha).
\end{Coqsrc}

@@

Lemma omega_second_AP :
  least_member   lt 
                  (fun alpha => 1 o< alpha /\ In AP alpha)
                  omega.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma AP_plus_closed (alpha beta gamma : Ord): 
     In AP alpha -> beta o< alpha -> gamma o< alpha -> beta + gamma o< alpha.

@@

Lemma phi0_mono (alpha beta : Ord) :
  alpha o< beta ->  phi0 alpha o< phi0 beta.

@@

Lemma is_limit_phi0 (alpha : Ord) :
  zero o< alpha ->  is_limit (phi0 alpha).

@@

Lemma phi0_le (alpha : Ord) : alpha o<= phi0 alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma lt_phi0 (alpha : Ord):
  alpha o< epsilon0 -> alpha o< phi0 alpha.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition Cr_fun : forall alpha : Ord,
       (forall beta : Ord, beta o< alpha -> Ensemble Ord) ->
        Ensemble Ord 
:= 
   fun (alpha :Ord)
        (Cr : forall beta, 
                beta o< alpha -> Ensemble Ord) 
        (x : Ord) => (
       (alpha = zero /\ AP x) \/
       (zero o< alpha /\
        forall beta (H:beta o< alpha),
          the_ordering_segment (Cr beta H) x /\ ord (Cr  beta H) x = x)).

@@

Definition sorted (l: cnf_t) :=
  LocallySorted (fun alpha beta => beta o<= alpha) l.

@@

\begin{Coqsrc}
Lemma cnf_lt_epsilon0 : 
 forall l alpha, 
   is_cnf_of alpha l ->  alpha o< epsilon0 ->
   Forall (fun beta =>  beta o< alpha) l.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma cnf_lt_epsilon0_iff : 
 forall l alpha, 
   is_cnf_of alpha l ->  
   (alpha o< epsilon0 <->  Forall (fun beta =>  beta o< alpha) l).
\end{Coqsrc}

@@

Theorem inject_mono (beta gamma : T1) :
  T1.lt  beta gamma -> 
  T1.nf beta -> T1.nf gamma -> 
  inject beta o< inject gamma.

@@

\begin{Coqsrc}
 Theorem inject_lt_epsilon0 (alpha : T1):
      inject alpha o< epsilon0.

### 2020-08-08 latex/chapter-termination.tex
specific infix for ordinal inequalities : first wave
corrections on pdf file

% Our formalisation of Cantor Normal Form will take two steps:
% 1 Definition of a strict order \texttt{<} on the type \texttt{T1}, 
% 2 Using \texttt{<} for characterizing terms in normal form.

@@

\begin{Coqsrc}
Definition LT := restrict nf lt.
Infix "<" := LT : t1_scope.

Definition LE := restrict nf le.
Notation "alpha <= beta" := (LE alpha beta) : t1_scope.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma LT_one : forall alpha, alpha < one -> alpha = zero.
\end{Coqsrc}

@@

Infix "<" := Lt : E0_scope.
Infix "<=" := Le : E0_scope.
\end{Coqsrc}

@@

Example e'2 : omega < omega + 6.
Proof. now compute. Qed.

@@

Example e'''2 : omega < omega * 6.
Proof. now compute. Qed.
\end{Coqsrc}

\begin{Coqsrc}
Lemma plus_not_monotonous : exists alpha beta gamma : T1,
    alpha < beta /\ alpha + gamma = beta + gamma.
Proof.
  exists 3, 5, omega;  now  compute.
Qed.

Lemma mult_not_monotonous :  exists alpha beta gamma : T1,
      alpha < beta /\ alpha * gamma = beta * gamma.
Proof.
  exists 3, 5, omega; now compute.
Qed.
\end{Coqsrc}

@@

  beta : T1
  n : nat
  gamma : T1
  IHbeta : nf beta -> Acc LT beta
  IHgamma : nf gamma -> Acc LT gamma
  H : nf (ocons beta n gamma)
  delta : T1
  Hdelta : delta < ocons beta n gamma
  ============================
  Acc LT delta
 \end{Coqanswer}

@@

  alpha : T1
  Aalpha : forall y : T1,  y < alpha -> Acc LT y
  IHalpha : forall y : T1,
       LT y alpha ->
       forall (n : nat) (beta : T1),
       nf (ocons y n beta) -> Acc LT (ocons y n beta)
  ============================
   forall (n : nat) (beta : T1),
   nf (ocons alpha n beta) -> Acc LT (ocons alpha n beta)
\end{Coqanswer}

@@

\begin{Coqsrc}
Lemma oplus_strict_mono_LT_l (alpha beta gamma : T1) :
  nf gamma   -> alpha  < beta ->
  alpha o+ gamma < beta o+ gamma.

Lemma oplus_strict_mono_LT_r (alpha beta gamma : T1) :
  nf alpha -> beta < gamma ->
  alpha o+ beta < alpha o+ gamma.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma S0_decr :
  forall s s', S0  s s' -> ms s' < ms s.
\end{Coqsrc}

\begin{Coqsrc}
Lemma R1_decr : forall h h',
                  R1 h h' -> m h' < m h.
\end{Coqsrc}

\begin{Coqsrc}
Lemma S1_decr n:
  forall s s', S1 n s s' -> ms s' <  ms s.
\end{Coqsrc}

\begin{Coqsrc}
Lemma R2_decr n : forall h h', R2 n h h' -> m h'  < m h.
\end{Coqsrc}

\begin{Coqsrc}
Lemma round_decr : forall h h', h -1-> h' -> m h' < m h.
Proof.
   destruct 1 as [n H]; destruct H. 
   -  now apply R1_decr.
   -  now apply R2_decr with n.
Qed.
\end{Coqsrc}

@@

Theorem every_battle_terminates: Termination.
Proof. 
  red; apply Inclusion.wf_incl with 
         (R2 := fun h h' =>  m h < m h').
   red; intros;  now apply round_decr.
   apply Inverse_Image.wf_inverse_image, T1_wf.
Qed.
\end{Coqsrc}

>>>>

% Our formalisation of Cantor Normal Form will take two steps:
% 1 Definition of a strict order \texttt{o<} on the type \texttt{T1}, 
% 2 Using \texttt{o<} for characterizing terms in normal form.

@@

\begin{Coqsrc}
Definition LT := restrict nf lt.
Infix "o<" := LT : t1_scope.

Definition LE := restrict nf le.
Notation "alpha o<= beta" := (LE alpha beta) : t1_scope.
\end{Coqsrc}

\begin{todo}
Infix notation for inequalities on ordinals have changed to \texttt{o<} and \texttt{o<=}.  This document is not updated yet, but the \coq{} scripts are correct.
\end{todo}

@@

\begin{Coqsrc}
Lemma LT_one : forall alpha, alpha o< one -> alpha = zero.
\end{Coqsrc}

@@

Infix "o<" := Lt : E0_scope.
Infix "o<=" := Le : E0_scope.
\end{Coqsrc}

\begin{todo}
Infix notation for inequalities on ordinals have changed to \texttt{o<} and \texttt{o<=}.  This document is not updated yet, but the \coq{} scripts are correct.
\end{todo}

@@

Example e'2 : omega o< omega + 6.
Proof. now compute. Qed.

@@

Example e'''2 : omega o< omega * 6.
Proof. now compute. Qed.
\end{Coqsrc}

\begin{Coqsrc}
Lemma plus_not_monotonous : exists alpha beta gamma : T1,
    alpha o< beta /\ alpha + gamma = beta + gamma.
Proof.
  exists 3, 5, omega;  now  compute.
Qed.

Lemma mult_not_monotonous :  exists alpha beta gamma : T1,
      alpha o< beta /\ alpha * gamma = beta * gamma.
Proof.
  exists 3, 5, omega; now compute.
Qed.
\end{Coqsrc}

@@

  beta : T1
  n : nat
  gamma : T1
  IHbeta : nf beta -> Acc LT beta
  IHgamma : nf gamma -> Acc LT gamma
  H : nf (ocons beta n gamma)
  delta : T1
  Hdelta : delta o< ocons beta n gamma
  ============================
  Acc LT delta
 \end{Coqanswer}

@@

  alpha : T1
  Aalpha : forall y : T1,  y o< alpha -> Acc LT y
  IHalpha : forall y : T1,
       LT y alpha ->
       forall (n : nat) (beta : T1),
       nf (ocons y n beta) -> Acc LT (ocons y n beta)
  ============================
   forall (n : nat) (beta : T1),
   nf (ocons alpha n beta) -> Acc LT (ocons alpha n beta)
\end{Coqanswer}

@@

\begin{Coqsrc}
Lemma oplus_strict_mono_LT_l (alpha beta gamma : T1) :
  nf gamma   -> alpha  o< beta ->
  alpha o+ gamma o< beta o+ gamma.

Lemma oplus_strict_mono_LT_r (alpha beta gamma : T1) :
  nf alpha -> beta o< gamma ->
  alpha o+ beta o< alpha o+ gamma.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma S0_decr :
  forall s s', S0  s s' -> ms s' o< ms s.
\end{Coqsrc}

\begin{Coqsrc}
Lemma R1_decr : forall h h',
                  R1 h h' -> m h' o< m h.
\end{Coqsrc}

\begin{Coqsrc}
Lemma S1_decr n:
  forall s s', S1 n s s' -> ms s' o<  ms s.
\end{Coqsrc}

\begin{Coqsrc}
Lemma R2_decr n : forall h h', R2 n h h' -> m h'  o< m h.
\end{Coqsrc}

\begin{Coqsrc}
Lemma round_decr : forall h h', h -1-> h' -> m h' o< m h.
Proof.
   destruct 1 as [n H]; destruct H. 
   -  now apply R1_decr.
   -  now apply R2_decr with n.
Qed.
\end{Coqsrc}

@@

Theorem every_battle_terminates: Termination.
Proof. 
  red; apply Inclusion.wf_incl with 
         (R2 := fun h h' =>  m h o< m h').
   red; intros;  now apply round_decr.
   apply Inverse_Image.wf_inverse_image, T1_wf.
Qed.
\end{Coqsrc}
